From physical principles to ???

### 2016-02-11 2015CompleteDerivation/completeDerivation.tex
Status and todos update
More on processes and outcomes

Before we are able to define how states should be captured mathematically, we need to better specify what we mean by physically distinguishable. Therefore we need to characterise what we mean by measurement, particularly because it has become such an overloaded term.

Consider the motion of a cannonball under inertial and gravitational forces. Light will be scattering off of it; as it lands the ground will be deformed and its temperature will rise. Those external physical process, which happen no matter what we do, can be used to distinguish the motion of the cannonball as their outcome are correlated. Therefore we can learn the position by looking at the reflected light, and learn the final kinetic momentum by looking at the deformation of the ground. In principle, any external process that has a correlation with the states under study can be used to perform a measurement, and any measurement is based on such a process. That is: for us a measurement is simply a physical process that we can use to distinguish states. There is no particular role played by the observer, except maybe create the device and forcing the system under study to interact with it.

This external process may be quite complicated: when a charged particle enters a calorimeter, a shower of particles is produced, photons are captured and are directed to photomultipliers, a current is read out, the current is then digitized, and so on. Some processes interfere with the system, they affect its dynamics, and other are destructive, the system no longer can be described by the original set of states. A bubble chamber is an example of the first (e.g. the magnetic field curves the motion of a charged particle); burning a substance to determine its caloric content is an example of the latter. In these cases, special care needs to make sure to link the outcomes with the original state, unaffected state. But in the end, the conceptual description remains the same.

We have a physical process that provides an outcome. This process has a set of potential outcomes. To each potential outcome we can associate a set of states. For example, if the electron follows this path, then its state is among the ones that has spin $1/2$. Since all measurements are finite precision, each outcome can be labelled by a finite set of digits, so the set of outcomes is countable. Note that different outcomes for the same process can overlap. For example, $4.12 \pm 0.05$ cm and $4.13 \pm 0.05$ cm are both legitimate possible outputs of the same measurement device. But since all states must be distinguishable, given two arbitrary states there must be a process precise enough to tell them apart. That is: the potential outcomes associated with the two states do not overlap. For example, $4.12 \pm 0.0005$ cm and $4.13 \pm 0.0005$ cm do not overlap anymore.

We can also combine two outcome of two different processes. For example, if one tells us spin is $+1/2$ and another that position is $4.13 \pm 0.05$ cm, then can say the state has both. That is, the state is inside both of the sets of states associated with the two outcomes: the intersection of two sets associated with an outcome is also a set associated with an outcome. We can make similar arguments about the union as we can always reduce the precision by combining different outcome of the same process.

@@

@@

%In practice, we use physical quantities to identify states. For example, the state for an ideal gas is given by pressure, volume and temperature. To clarify whether we are referring to a particular value of such quantity, or the quantity itself, we introduce the following terminology:
%\begin{center}
%    \begin{tabular}{ | p{2.5cm} | p{5.5cm} | }
%    \hline
%    State variable & A quantity (discrete or continuous) that must be specified to identify a state (e.g. position). \\ \hline
%    Label & A particular value for a state variable (e.g. position = $5m$). \\ \hline
%    Label range & A set of possible values for a state variable (e.g. position = $[4.5m, 5.5m]$). \\
%    \hline
%    \end{tabular}
%\end{center}

To identify and name states, we will typically use a set of quantities, typically numbers. For example, the orbital of an electron in a hydrogen item is identified by the quantum numbers $n$, $l$, $m$ and $s$. We call each of these quantities \emph{state variable} and \emph{state vector} a set of state variables that can identify one and only one state. We avoid the term coordinates, even though that's what they mathematically are, as it will create confusion with space coordinates. We also avoid the term observable or measurable, as not all state variables map directly to the outcome of an experiment (e.g. conjugate momentum as it is not gauge invariant). It's clear, though, that even in those cases there must exist, at least in principle, a way to distinguish between the possible values of the state variable (e.g once we fix the vector potential, conjugate momentum can be measured via the kinetic momentum). Therefore we can assume that all state variables are in the end physically distinguishable, yet we do not assume all state variables have equal standing in physical meaning. Conversely, some measurable quantities are not directly state variables: they may not be defined, as states are, at constant time (e.g. velocity) or may depend on the process at hand (e.g. acceleration). Yet, once a process is fully defined (e.g. a Hamiltonian is given), we may have a relationship between those observables and state variables. The purpose of state variables is just to label states, which are physically distinguishable and defined at a particular time. Therefore the state variable will be physically distinguishable and defined at constant time. But there is nothing telling us that they must be directly physically meaningful or that their relationship to measurable quantities is the same if we change physical processes.

@@

The second approach is to transport the state variables. The idea is to keep the connection to the initial state by labelling the future state, instead by the future values of $q^i$, by the original one. So we introduce a new set of variables for which $\hat{q}^i(\hat{\mathcal{s}})=q^i(\mathcal{s})$, which we can always do as the evolution is deterministic and reversible. The level sets of $\hat{q}^i$ are the evolved level sets of $q^i$, which allows us to study how groups of states evolve in time. Moreover, once we identify an invariant for $q^i$ this is automatically satisfied by $\hat{q}^i$. Given that the value of the transported state variables does not change during evolution, and that it is unique for each initial state, transported state variable provide a way to label the trajectories themselves. While the numeric value of the transported state variables remains the same, the state variable (and its physical meaning) changes. For example, pressure and temperature will be combined depending on the particular process.

@@

\begin{justification}
TODO: set of state variables locally define a chart; inherit topology from charts (small change in state makes small change in state variable, makes small changes in outcomes?)
\end{justification}

@@

@@

>>>>

Before we are able to define how states should be captured mathematically, we need to better specify what we mean by physically distinguishable. Therefore we need to characterize what we mean by measurement, particularly because it has become such an overloaded term.

Consider the motion of a cannonball under inertial and gravitational forces. Light will be scattering off of it; as it lands the ground will be deformed and its temperature will rise. Those external physical processes, which happen no matter what we do, can be used to distinguish the motion of the cannonball as their outcomes are correlated. Therefore we can learn the position by looking at the reflected light, and learn the final kinetic momentum by looking at the deformation of the ground. In principle, any external process that has a correlation with the states under study can be used to perform a measurement, and any measurement is based on such a process. That is: for us a measurement is simply a physical process that we can use to distinguish states. There is no particular role played by the observer, except maybe create the device and forcing the system under study to interact with it.

This external process may be quite complicated: when a particle enters a calorimeter, a shower of particles is produced, photons are captured and are directed to photomultipliers, a current is read out, the current is then digitized, and so on. Some processes interfere with the system, they affect its dynamics, and other are destructive, the system no longer can be described by the original set of states. A tracking chamber is an example of the first (the magnetic field curves the motion of a charged particle); burning a substance to determine its caloric content is an example of the latter. In these cases, special care is needed to make sure to link the outcomes with the original state, unaffected state. But in the end, the conceptual description remains the same.

%TODO: It's not about precision! Discrete variables are known to infinite precision! Is about information.

We have a physical process that provides an outcome. This process has a set of potential outcomes. To each potential outcome we can associate a set of states. For example, if the electron follows a certain path, then its state is among the ones that has spin $+1/2$. Since all measurements are finite precision, each outcome can be labeled by a finite set of digits, so the set of outcomes is countable. Note that different outcomes for the same process can overlap. For example, $4.12 \pm 0.05$ cm and $4.13 \pm 0.05$ cm are both legitimate possible outputs of the same measurement device. But since all states must be distinguishable, given two arbitrary states there must be a process precise enough to tell them apart. That is: the potential outcomes associated with the two states do not overlap. For example, $4.12 \pm 0.0005$ cm and $4.13 \pm 0.0005$ cm do not overlap anymore.

%TODO If we start "combining processes" then is it really _a_ process?

We can also combines two outcome of two different processes. For example, if one tells us spin is $+1/2$ and another that position is $4.13 \pm 0.05$ cm, then we can say the state has both. That is, the state is inside both of the sets of states associated with the two outcomes: the intersection of two sets associated with an outcome is also a set associated with an outcome. We can make similar arguments about the union as we can always reduce the precision by combining different outcome of the same process.

@@

With our state space defined, deterministic and reversible evolution corresponds to a bijective map between initial and final states.

@@

To identify and name states we will use a set of quantities, typically numbers. For example, the orbital of an electron in a hydrogen item is identified by the quantum numbers $n$, $l$, $m$ and $s$. We call each of these quantities \emph{state variable} and \emph{state vector} a set of state variables that can identify one and only one state. We avoid the term coordinates, even though that's what they mathematically are, as it will create confusion with space coordinates. We also avoid the term observable or measurable, as not all state variables map directly to the outcome of an experiment (e.g. conjugate momentum as it is not gauge invariant). It's clear, though, that even in those cases there must exist, at least in principle, a way to distinguish between the possible values of the state variable (e.g once we fix the vector potential, conjugate momentum can be measured via the kinetic momentum). Therefore we can assume that all state variables are in the end physically distinguishable, yet we do not assume all state variables have equal standing in physical meaning. Conversely, some measurable quantities are not directly state variables: they may not be defined, as states are, at constant time (e.g. velocity) or may depend on the process at hand (e.g. acceleration). Yet, once a process is fully defined (e.g. a Hamiltonian is given), we may have a relationship between those observables and state variables. The purpose of state variables is just to label states, which are physically distinguishable and defined at a particular time. Therefore the state variable will be physically distinguishable and defined at constant time. But there is nothing telling us that they must be directly physically meaningful or that their relationship to measurable quantities is the same if we change physical processes.

@@

The second approach is to transport the state variables. The idea is to keep the connection to the initial state by labeling the future state, instead by the future values of $q^i$, by the original one. So we introduce a new set of variables for which $\hat{q}^i(\hat{\mathcal{s}})=q^i(\mathcal{s})$, which we can always do as the evolution is deterministic and reversible. The level sets of $\hat{q}^i$ are the evolved level sets of $q^i$, which allows us to study how groups of states evolve in time. Moreover, once we identify an invariant for $q^i$ this is automatically satisfied by $\hat{q}^i$. Given that the value of the transported state variables does not change during evolution, and that it is unique for each initial state, transported state variable provide a way to label the trajectories themselves. While the numeric value of the transported state variables remains the same, the state variable (and its physical meaning) changes. For example, pressure and temperature will be combined depending on the particular process.

@@

\begin{justification}
TODO: set of state variables locally define a chart, isomorphic to $\mathbbm{R}$.
\end{justification}

@@

%Status: conceptual work is done, need to write down

%TODO: need more examples within explanation to make more intuitive

@@

%TODO: move second picture to QM
